Open Clusters Characterization in Gaia DR2 Using ML Algorithms

### 2020-11-09 thesis/thesis.tex
thesis: Add structure comments
thesis: Update structure

@@

\section{Aims}

The goal of this work is to \emph{achieve a viable model for open cluster characterization based on machine learning techniques}
such as clustering algorithms and artificial neural networks.
This model must be a non-supervised non-parameterized algorithm valid for any selected OC.

The OpenClust catalogue \cite{dias2002new} has been used to make a preselection of sky regions for the Gaia DR2 data download.
This download is not limited to the sized registered in the catalogue. Instead, a wider region is downloaded for each cluster to include
stars from outside the cluster and take them into account when training the model so the model will be able to take apart those outsiders
when characterizing a non known cluster.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{../figures/openclust_catalogue.pdf}
  \caption{OpenClust Catalogue Distribution}
\end{figure}

A later filter will be applied to select only those OCs that fulfil one or more of the following criteria:

\begin{itemize}
  \item Cluster diameter above 25.0 arcmin
  \item Number of stars in the selected region above 40,000 stars
\end{itemize}

The final downloaded set covers nearly 60 million stars, which is significantly smaller than the whole Gaia DR2 dataset which contains
information for around 1,600 million stars.
This restriction responds to a commitment to the project delivery dates and the available computing power,
and to having a good basis for the correct model training. As an extension of this work, the trained model could be applied to a region
not covered by this preselection for a later understanding of the results.

The intention is to get a reliable and efficient non-parameterized neither supervised model over those parameterized or supervised ones.
To measure the reliability of the resulting model comparisons with other VO tools will be made.

The role for the ANN will be validating cluster members based on dynamincs and photometric data.

\begin{itemize}
  \item OCs and stars \emph{right ascension}: $\alpha$ and \emph{declination}: $\delta$
  \item OCs \emph{estimated sizes}: $r$
  \item \emph{Proper motion}: $\mu_{\alpha}$, $\mu_{\delta}$ and \emph{parallax}: $\varpi$
  \item \emph{Photometric}: $E(B-R)$, $G_{mag}$
  \item \emph{Associated errors} will be taken into account to estimate prediction errors
\end{itemize}

Initially, no filter will be applied over parallax neither magnitude values to the selected objects from Gaia DR2.

@@

\chapter{Requirements Description}

\chapter{Work Development}

@@

>>>>

%% TODO: 1.1 Motivation

% ¿Cuál es el problema a tratar?
% ¿Cuáles crees que son las causas?
% ¿Por qué es relevante el problema?

%% TODO: 1.2 Planteamiento

% ¿Cómo se podría solucionar el problema?
% ¿Qué es lo que se propone?
% Descripción del objetivo del trabajo en términos generales

%% TODO: 1.3 Estructura de la memoria

% Descripción breve de lo que se va a contar en cada uno de los capítulos siguientes.

@@

@@

%% TODO: Concluir con una última sección de resumen de conclusiones, resumiendo las principales averiguaciones del estudio del dominio
% y cómo van a afectar al desarrollo específico del trabajo.

\chapter{Aims}

\section{General}

The main aim of this thesis is to \emph{build an unsupervised clustering model for open cluster characterization}.
This model must be \emph{non-supervised non-parameterized}, and has to be valid for a wide range of open clusters.

To measure the reliability of the resulting model comparisons with other VO tools will be made.

The role for the ANN will be validating cluster members based on dynamincs data.

\begin{itemize}
  \item OCs and stars \emph{right ascension}: $\alpha$ and \emph{declination}: $\delta$
  \item OCs \emph{estimated sizes}: $r$
  \item \emph{Proper motion}: $\mu_{\alpha}$, $\mu_{\delta}$ and \emph{parallax}: $\varpi$
  \item \emph{Associated errors} will be taken into account to estimate prediction errors
\end{itemize}

\section{Specific}

To achive this goal, the aim will be divided into the following milestones:

\begin{itemize}
  \item To recover data from Gaia DR2 database. This data will be taken as source for the machine learning model to characterize open clusters
  by grouping stars into clusters.
  \item To research unsupervised clustering algorithms suitable for grouping stars previously recovered.
  \item To select and implement unsupervised clustering algorithms from the previous study.
  \item To apply implemented tools to different datasets from first point to find OCs.
  \item To validate found OCs by determining their accuracy and precission.
\end{itemize}

\chapter{Method}

In this chapter we address the different steps taken to perform the creation of the unsupervised clustering model for open cluster characterization.

The development has been made with the Python programming language. Other auxiliary tools such as \emph{Docker}, \emph{PostgreSQL}, \emph{Jupyter Notebooks},
and \emph{Scikit-Learn} and \emph{Keras} frameworks have been used too.

\section{Data Mining}

As mentioned before, Gaia DR2 is the chosen data source for this work. Thus, the first step is to download the required data from Gaia DB
and store it in a custom database for later access.

Due to the large amount of data inside Gaia database, a complete download is not viable neither useful.
In order to squize the size of the dataset to be downloaded, the OpenClust catalogue \cite{dias2002new} (Figure \ref{fig:OpenClustComplete})
has been used to make a preselection of sky regions.

This download is not limited to the size registered in the catalogue. Instead, a wider region is downloaded for each cluster to include
stars outside the cluster and take them into account when training the model so the model will be able to take apart those outsiders
when characterizing the cluster.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{../figures/openclust_catalogue.pdf}
  \caption{OpenClust Catalogue Distribution}
  \label{fig:OpenClustComplete}
\end{figure}

The downloaded dataset covers nearly 114 million stars (42GB of compressed data), that is significantly smaller than the whole Gaia DR2 dataset, which contains
information for around 1,600 million stars, but stils being too large.

Since not all the downloaded clusters are good enough, the following filters are applied to discard some clusters which don't have enough stars or contain too
many \emph{null} values:

\begin{itemize}
  \item Cluster diameter above 25.0 arcmin
  \item Parallax absolute value greater than 0.0
  \item Number of stars in the selected region above 40,000 stars
\end{itemize}

There is no upper limit to the number of stars inside a cluster, although for the sake of this work, smaller cluster will be prefered as
a commitment to the project delivery dates and the available computing power. As an extension of this work, the trained model could be applied to a region
not covered by this preselection for a later understanding of the results.

%% TODO: Las tecnologías empleadas para el desarrollo de las herramientas,
% la organización del piloto, cómo ha transcurrido el experimento

%% TODO: Añadir nuevo capítulo - Descripción de los resultados

% Aquí se detallan los resultados obtenidos, con tablas de resumen, gráficas de resultados, identificación de datos relevantes, etc.
% Es una exposición objetiva, sin valorar los resultados ni justificarlos.

\chapter{Discussion}

% Se aporta la discusión de los resultados presentados en el capítulo anterior. En este capítulo se puede discutir la relevancia de
% los resultados, presentar posibles explicaciones apra los datos anómalos y resaltar aquellos datos que sean parcitualarmente relevantes
% para el análisis del experimento.

@@

\section{Conclusions}

% Resumen final del trabajo y debe servir para informar del alcance y relevancia de la aportación.
% Suele estructurarse empezando con un resuemn del problema tratado, de cómo se ha abordado y de por qué la solución sería valída.

% Es recomendable que incluya también un resumen de las contribuciones del trabajo, en el que se relacionen las contribuciones y los
% resultado obtenidos con ls objetivos panteados para el trabajo, discutiendo hasta qué punto has conseguido resolver los objetivos planteados.

\section{Further Research}

% Debería señalar las perspectivas de futuro que abre el trabajo desarrollado para el campo de estudio definido. En el fondo,
% debe justificar de qué modo puede emplearse la aportación desarrollada y en qué campos.
